---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%"
)
options(width = 100)
```

# panglm

Fast generalized linear models for panel data: pooled, fixed-effects
("within"), and random-effects estimators for gaussian, poisson, binomial,
and negative binomial outcomes, backed by an Rcpp/RcppArmadillo/RcppParallel
numerical core. Every estimator here is validated against an external
reference implementation (`plm`, `pglm`, `fixest`, `MASS::glm.nb`,
`survival::clogit`, `pscl::hurdle`, or `glmmTMB`), not just checked for
"it runs" -- see the Validation section below and `vignette("panglm-methods")`.

## Installation

```r
# install.packages("remotes")
remotes::install_github("ielbadisy/panglm")
```

## Fixed effects, exactly matching `plm`

```{r within-gaussian}
library(panglm)
data(Grunfeld, package = "plm")

f_panglm <- panglm(inv ~ value + capital, data = Grunfeld,
                    index = c("firm", "year"), model = "within", family = "gaussian")
f_plm    <- plm::plm(inv ~ value + capital, data = Grunfeld,
                      index = c("firm", "year"), model = "within")

summary(f_panglm)
```

```{r within-gaussian-diff}
cbind(panglm = coef(f_panglm), plm = coef(f_plm), diff = coef(f_panglm) - coef(f_plm))
```

```{r coef-plot, fig.width=6, fig.height=3.2}
est <- rbind(
  data.frame(term = names(coef(f_panglm)), estimate = coef(f_panglm),
             se = sqrt(diag(vcov(f_panglm))), fit = "panglm"),
  data.frame(term = names(coef(f_plm)), estimate = coef(f_plm),
             se = sqrt(diag(vcov(f_plm))), fit = "plm")
)
ggplot2::ggplot(est, ggplot2::aes(x = estimate, y = term, color = fit)) +
  ggplot2::geom_pointrange(
    ggplot2::aes(xmin = estimate - 1.96 * se, xmax = estimate + 1.96 * se),
    position = ggplot2::position_dodge(width = 0.4), fatten = 3
  ) +
  ggplot2::scale_color_manual(values = c(panglm = "#2a78d6", plm = "#eb6834")) +
  ggplot2::labs(title = "Fixed-effects coefficients: panglm vs. plm (Grunfeld)",
                x = "Estimate (95% CI)", y = NULL, color = NULL) +
  ggplot2::theme_minimal(base_size = 11)
```

The two estimates are indistinguishable, as they should be: `panglm`'s
`model = "within"` is an exact, closed-form demeaning estimator, not an
approximation of `plm`'s.

## Count panels: fixed-effects Poisson and negative binomial

```{r within-poisson}
set.seed(1)
Grunfeld$count <- rpois(nrow(Grunfeld), lambda = exp(0.5 + 0.0001 * Grunfeld$value))

f_pois <- panglm(count ~ value + capital, data = Grunfeld,
                  index = c("firm", "year"), model = "within", family = "poisson")
coef(f_pois)
```

```{r within-negbin}
u <- rep(rgamma(10, shape = 5, rate = 5), each = 20)
Grunfeld$count_nb <- rpois(nrow(Grunfeld), lambda = u * exp(1 + 0.0002 * Grunfeld$value))

f_nb <- panglm(count_nb ~ value + capital, data = Grunfeld,
               index = c("firm", "year"), model = "within", family = "negbin")
coef(f_nb)
fixest::fenegbin(count_nb ~ value + capital | firm, data = Grunfeld) |> coef()
```

`family = "negbin"` uses the Allison-Waterman (2002) unconditional
dummy-variable estimator and matches `fixest::fenegbin()` exactly, including
`effect = "twoways"` (individual + time fixed effects); `family = "poisson"`
uses the conditional (concentrated) MLE (Hausman, Hall & Griliches 1984),
matching `pglm(model = "within", family = poisson)`.

## Structural zeros: a fixed-effects hurdle model

Ordinary Poisson/NB can't distinguish a *structural* zero (a group that
never has the event) from an *ordinary* sampling zero at low mean -- it
just shows up as extreme overdispersion regardless of family.
`panglm_hurdle()` decomposes the two:

```{r hurdle}
set.seed(9)
N <- 60; Tt <- 8
d <- data.frame(id = rep(1:N, each = Tt), time = rep(1:Tt, N))
d$x1  <- rnorm(N * Tt)
a_i   <- rep(rnorm(N, sd = 1.5), each = Tt)
mu    <- exp(0.3 + 0.5 * d$x1 + a_i)
p0    <- plogis(1 - 1.2 * a_i)
zero  <- rbinom(N * Tt, 1, p0) == 1
d$y   <- ifelse(zero, 0, rpois(N * Tt, mu))

fit <- panglm_hurdle(y ~ x1, data = d, index = c("id", "time"))
fit
```

The count part (zero-truncated Poisson, fit only on `y > 0`) recovers the
true slope of `0.5`; the zero part is fit via the same exact conditional
logistic regression `panglm(..., family = "binomial")` already uses.

## Standard errors: classical, HC1, or cluster-robust

```{r vcov}
sqrt(diag(vcov(f_panglm, type = "classical")))
sqrt(diag(vcov(f_panglm, type = "cluster")))
```

## What's estimated, and against what

| `model` | `family` | Method | Validated against |
|---|---|---|---|
| `pooling` | gaussian/poisson/binomial | IRLS | `stats::glm()` |
| `pooling` | negbin | theta-profiled IRLS | `MASS::glm.nb()` |
| `within` | gaussian | exact demeaning (one-way or `twoways`) | `plm(model="within")` |
| `within` | poisson | conditional MLE (one-way or `twoways`) | `pglm()`, `fixest::feglm()` |
| `within` | negbin | Allison-Waterman dummy-variable (one-way or `twoways`) | `fixest::fenegbin()` |
| `within` | binomial | exact conditional logit (Chamberlain 1980) | `survival::clogit(method="exact")` |
| `random` | gaussian | Swamy-Arora | `plm(model="random")` |
| `random` | poisson | Poisson-Gamma marginal | `pglm(model="random")` |
| `random` | negbin | 2-parameter beta-negative-binomial marginal | `pglm()`'s `lnl.negbin.R` |
| `random` | binomial | Gauss-Hermite quadrature | `glmmTMB` (approximate, different mixing distribution) |
| -- | -- | `panglm_hurdle()`: conditional logit + zero-truncated Poisson | `pscl::hurdle()` |

Also included: `panglm_hausman()` (FE-vs-RE specification test),
`panglm_dispersiontest()` (Pearson chi-squared/df overdispersion test),
`confint.panglm()`, and `tidy()`/`glance()` methods for `broom`/`modelsummary`
pipelines.

Not yet covered: `effect = "twoways"` for binomial (no general closed-form
two-way conditional logit exists), between effects, ordinal/tobit families,
and a zero-truncated NB2 count part for `panglm_hurdle()`. See `NEWS.md`.

## Validation

Every estimator above is checked against its reference implementation in
`tests/testthat/`, not just smoke-tested:

```{r test-count}
res <- testthat::test_local(reporter = testthat::SilentReporter$new())
df  <- as.data.frame(res)
c(pass = sum(df$passed), fail = sum(df$failed), warn = sum(df$warning), skip = sum(df$skipped))
```

See `vignette("panglm-methods")` for the full walkthrough (33+ executable
validation chunks) and `NEWS.md` for what changed and when.
