---
title: "panglm: models, likelihoods, and validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{panglm: models, likelihoods, and validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = requireNamespace("plm", quietly = TRUE) && requireNamespace("pglm", quietly = TRUE) &&
    requireNamespace("survival", quietly = TRUE)
)
```

## Why this vignette exists

`panglm` re-implements a specific, checkable set of panel-GLM likelihoods in
Rcpp/RcppArmadillo/RcppParallel instead of R-level `maxLik` optimization
(`pglm`) or symbolic differentiation frameworks (`glmmTMB`). Every estimator
below is validated against a reference implementation in the package's test
suite (`tests/testthat/`), not just benchmarked for speed. This document
states exactly which likelihood each `model`/`family` combination solves, so
results can be trusted rather than treated as a black box.

```{r setup}
library(panglm)
library(plm)
library(pglm)
library(survival)
data(Grunfeld, package = "plm")
data(UnionWage, package = "pglm")
```

## 1. Pooled models

`model = "pooling"` fits a standard GLM (gaussian/poisson/binomial, any of
the canonical links) via IRLS in C++, ignoring the panel structure entirely
-- this is the `glm()` model, just faster. `family = "negbin"` fits a
genuine NB2 model (alternating IRLS-for-beta / Newton-for-theta, the same
algorithm `MASS::glm.nb()` uses), not an alias for Poisson.

```{r pooled}
f_pool <- panglm(inv ~ value + capital, data = Grunfeld,
                  index = c("firm", "year"), model = "pooling", family = "gaussian")
g_pool <- glm(inv ~ value + capital, data = Grunfeld, family = gaussian())
rbind(panglm = coef(f_pool), glm = coef(g_pool))
```

```{r pooled-negbin, eval = requireNamespace("MASS", quietly = TRUE)}
set.seed(1)
Grunfeld$count <- rnbinom(nrow(Grunfeld), size = 3, mu = exp(0.5 + 0.0002 * Grunfeld$value))
f_nb <- panglm(count ~ value + capital, data = Grunfeld,
               index = c("firm", "year"), model = "pooling", family = "negbin")
g_nb <- MASS::glm.nb(count ~ value + capital, data = Grunfeld)
rbind(panglm = coef(f_nb), glm.nb = coef(g_nb))
```

## 2. Fixed effects ("within")

- **gaussian**: exact within (group-demeaned) OLS, identical to `plm(model =
  "within")`.
- **poisson**: the *conditional* (concentrated) Poisson likelihood that
  profiles out the individual fixed effect in closed form -- this is exactly
  `pglm`'s `model = "within", family = poisson` estimator (Hausman, Hall &
  Griliches 1984), not an approximation.
- **negbin**: the Allison & Waterman (2002) unconditional dummy-variable
  estimator -- one intercept dummy per individual, estimated jointly with
  the covariate slopes and NB2 dispersion by exact MLE. Unlike the
  conditional (Hausman-Hall-Griliches 1984) FE-NB estimator, this doesn't
  suffer the "fake fixed effects" critique (Guimaraes 2008): the
  incidental-parameters bias that would contaminate a fixed-effects
  *logit* dummy-variable estimator does not carry over to NB2 (Allison &
  Waterman 2002). Matches `fixest::fenegbin()` exactly. Tractable for
  small-to-moderate N (the design matrix grows with N); not intended for
  very large panels.
- **binomial**: the exact conditional logistic regression that profiles out
  the individual fixed effect by conditioning on each unit's total number of
  successes (Chamberlain 1980) -- the same estimator as
  `survival::clogit(method = "exact")`. Units with no within-group outcome
  variation (all-0 or all-1) carry no information and are dropped, with a
  message reporting how many.

```{r within}
f_fe <- panglm(inv ~ value + capital, data = Grunfeld,
                index = c("firm", "year"), model = "within", family = "gaussian")
p_fe <- plm(inv ~ value + capital, data = Grunfeld, index = c("firm", "year"), model = "within")
rbind(panglm = coef(f_fe), plm = coef(p_fe))
```

```{r within-negbin, eval = requireNamespace("fixest", quietly = TRUE)}
u <- rep(rnorm(10, sd = 0.5), each = 20)
Grunfeld$count_fe <- rnbinom(nrow(Grunfeld), size = 4, mu = exp(1 + 0.0002 * Grunfeld$value + u))
f_fenb <- panglm(count_fe ~ value + capital, data = Grunfeld,
                  index = c("firm", "year"), model = "within", family = "negbin")
fx_fenb <- fixest::fenegbin(count_fe ~ value + capital | firm, data = Grunfeld, notes = FALSE)
rbind(panglm = coef(f_fenb), fixest = coef(fx_fenb))
```

```{r within-binomial}
UnionWage$union01 <- as.numeric(UnionWage$union) - 1
f_clogit <- panglm(union ~ wage + exper + rural, data = UnionWage,
                    index = c("id", "year"), model = "within", family = "binomial")
cl <- survival::clogit(union01 ~ wage + exper + rural + strata(id), data = UnionWage, method = "exact")
rbind(panglm = coef(f_clogit), clogit = coef(cl))
```

### Two-way fixed effects

`effect = "twoways"` adds a time fixed effect for **gaussian** (exact
alternating-projections demeaning, Guimaraes & Portugal 2010) and
**poisson** (outer-IRLS / inner-weighted-FWL -- the algorithm behind
`fixest::feglm()`'s multi-way fixed effects); not yet for negbin/binomial.

```{r twoways}
f_tw <- panglm(inv ~ value + capital, data = Grunfeld, index = c("firm", "year"),
                model = "within", family = "gaussian", effect = "twoways")
p_tw <- plm(inv ~ value + capital, data = Grunfeld, index = c("firm", "year"),
            model = "within", effect = "twoways")
rbind(panglm = coef(f_tw), plm = coef(p_tw))
```

```{r twoways-poisson, eval = requireNamespace("fixest", quietly = TRUE)}
set.seed(1)
Grunfeld$count2 <- rpois(nrow(Grunfeld), lambda = exp(0.5 + 0.0001 * Grunfeld$value))
f_twp <- panglm(count2 ~ value + capital, data = Grunfeld, index = c("firm", "year"),
                 model = "within", family = "poisson", effect = "twoways")
fx_twp <- fixest::feglm(count2 ~ value + capital | firm + year, data = Grunfeld, family = poisson())
rbind(panglm = coef(f_twp), fixest = coef(fx_twp))
```

## 3. Random effects

- **gaussian**: the Swamy-Arora quasi-demeaning (feasible GLS) estimator,
  identical to `plm(model = "random")`'s default.
- **poisson**: the Poisson-Gamma (negative-binomial marginal) closed-form
  likelihood -- `pglm`'s `model = "random", family = poisson` estimator. The
  fitted dispersion parameter is the Gamma frailty shape. Seeded from the
  pooled-Poisson MLE plus a method-of-moments dispersion estimate, with
  Levenberg-Marquardt/ridge damping (and a warning) instead of erroring if
  the Hessian is ill-conditioned -- verified stable on heavily
  zero-inflated, extremely skewed panels.
- **negbin**: a *different* closed-form mixture from `poisson` above -- a
  2-parameter beta-negative-binomial marginal (`pglm`'s `lnl.negbin.R`
  "random" branch exactly, with independent shape parameters `a`/`b` rather
  than Poisson-Gamma's single dispersion).
- **binomial**: the individual random intercept is integrated out via
  Gauss-Hermite quadrature (`R` nodes, default 21), matching `pglm`'s
  `model = "random", family = binomial(...)`.

```{r random-poisson}
set.seed(42)
Grunfeld$u <- rep(rgamma(10, shape = 5, rate = 5), each = 20)
Grunfeld$count <- rpois(nrow(Grunfeld), lambda = Grunfeld$u * exp(1 + 0.0002 * Grunfeld$value))
f_re <- panglm(count ~ value + capital, data = Grunfeld,
                index = c("firm", "year"), model = "random", family = "poisson")
p_re <- pglm(count ~ value + capital, data = Grunfeld, index = c("firm", "year"),
             model = "random", family = poisson, method = "bfgs", print.level = 0, R = 5)
rbind(panglm = coef(f_re), pglm = coef(p_re)[1:3])
```

```{r random-negbin}
set.seed(3)
Grunfeld$u2 <- rep(rgamma(10, shape = 6, rate = 6), each = 20)
Grunfeld$count3 <- rnbinom(nrow(Grunfeld), size = 8, mu = Grunfeld$u2 * exp(1 + 0.0002 * Grunfeld$value))
f_renb <- panglm(count3 ~ value + capital, data = Grunfeld,
                  index = c("firm", "year"), model = "random", family = "negbin")
coef(f_renb)
c(a = f_renb$negbin_a, b = f_renb$negbin_b)
```

```{r random-binomial}
f_bin <- panglm(union ~ wage + exper + rural, data = UnionWage,
                 index = c("id", "year"), model = "random", family = "binomial", R = 15)
p_bin <- pglm(union ~ wage + exper + rural, UnionWage, index = c("id", "year"),
              family = binomial("logit"), model = "random", method = "bfgs",
              print.level = 0, R = 15)
rbind(panglm = coef(f_bin), pglm = coef(p_bin)[1:4])
```

## 4. Standard errors

Model-based (classical) SEs are always available. `vcov(fit, type = "HC1")`
and `vcov(fit, type = "cluster")` add heteroskedasticity-robust and
cluster-robust (Stata-style CR1, clustered by the panel individual by
default) sandwich covariances for `model = "pooling"`/`"within"`:

```{r robust}
sqrt(diag(vcov(f_fe)))                       # classical
sqrt(diag(vcov(f_fe, type = "cluster")))     # cluster-robust
```

Random-effects models report model-based (information-matrix) SEs only --
the individual-level correlation is already accounted for in the
likelihood, matching the convention used by `lme4`/`glmmTMB`.

You can also request a non-classical vcov directly when fitting:
`panglm(..., vcov = "cluster")`.

## 4b. Confidence intervals and overdispersion

`confint()` gives Wald intervals from whichever vcov the fit currently
carries. `panglm_dispersiontest()` is a quick Pearson chi-squared/df
overdispersion diagnostic for poisson/negbin fits with available
`fitted.values` (values well above 1 suggest a negative binomial
specification instead):

```{r confint-dispersion}
confint(f_fe)
panglm_dispersiontest(f_nb)
```

## 5. Hausman test

`panglm_hausman()` compares a `"within"` and a `"random"` fit of the same
formula/family, replicating `plm::phtest()`:

```{r hausman}
panglm_hausman(f_fe, panglm(inv ~ value + capital, data = Grunfeld,
                             index = c("firm", "year"), model = "random", family = "gaussian"))
```

## 6. broom / modelsummary

`tidy()` and `glance()` methods are registered for the `generics` package's
generics, so `panglm` fits work directly with `broom`- and
`modelsummary`-based table pipelines:

```{r broom}
broom::tidy(f_pool)
```

## What's not (yet) covered

- Between and random-effects `ordinal`/`tobit` models (`pglm` has these via
  Gauss-Hermite quadrature; `panglm` doesn't yet).
- `effect = "twoways"` for binomial (no general closed-form two-way
  conditional logit exists; gaussian, poisson, and negbin are covered).
- Cluster/HC1-robust vcov for fixed-effects binomial (classical,
  Hessian-based SEs only for now; negbin is covered via the full
  covariate+dummy score, see `?vcov.panglm`).
- `panglm_hurdle()`'s count part is zero-truncated Poisson only; a
  zero-truncated NB2 count part is not yet implemented.
