---
title: "panglm: models, likelihoods, and validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{panglm: models, likelihoods, and validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = requireNamespace("plm", quietly = TRUE) && requireNamespace("pglm", quietly = TRUE) &&
    requireNamespace("survival", quietly = TRUE)
)
```

## Why this vignette exists

`panglm` re-implements a specific, checkable set of panel-GLM likelihoods in
Rcpp/RcppArmadillo/RcppParallel instead of R-level `maxLik` optimization
(`pglm`) or symbolic differentiation frameworks (`glmmTMB`). Every estimator
below is validated against a reference implementation in the package's test
suite (`tests/testthat/`), not just benchmarked for speed. This document
states exactly which likelihood each `model`/`family` combination solves, so
results can be trusted rather than treated as a black box.

```{r setup}
library(panglm)
library(plm)
library(pglm)
library(survival)
data(Grunfeld, package = "plm")
data(UnionWage, package = "pglm")
```

## 1. Pooled models

`model = "pooling"` fits a standard GLM (gaussian/poisson/binomial, any of
the canonical links) via IRLS in C++, ignoring the panel structure entirely
-- this is the `glm()` model, just faster.

```{r pooled}
f_pool <- panglm(inv ~ value + capital, data = Grunfeld,
                  index = c("firm", "year"), model = "pooling", family = "gaussian")
g_pool <- glm(inv ~ value + capital, data = Grunfeld, family = gaussian())
rbind(panglm = coef(f_pool), glm = coef(g_pool))
```

## 2. Fixed effects ("within")

- **gaussian**: exact within (group-demeaned) OLS, identical to `plm(model =
  "within")`.
- **poisson**: the *conditional* (concentrated) Poisson likelihood that
  profiles out the individual fixed effect in closed form -- this is exactly
  `pglm`'s `model = "within", family = poisson` estimator (Hausman, Hall &
  Griliches 1984), not an approximation.
- **binomial**: the exact conditional logistic regression that profiles out
  the individual fixed effect by conditioning on each unit's total number of
  successes (Chamberlain 1980) -- the same estimator as
  `survival::clogit(method = "exact")`. Units with no within-group outcome
  variation (all-0 or all-1) carry no information and are dropped, with a
  message reporting how many.

```{r within}
f_fe <- panglm(inv ~ value + capital, data = Grunfeld,
                index = c("firm", "year"), model = "within", family = "gaussian")
p_fe <- plm(inv ~ value + capital, data = Grunfeld, index = c("firm", "year"), model = "within")
rbind(panglm = coef(f_fe), plm = coef(p_fe))
```

```{r within-binomial}
UnionWage$union01 <- as.numeric(UnionWage$union) - 1
f_clogit <- panglm(union ~ wage + exper + rural, data = UnionWage,
                    index = c("id", "year"), model = "within", family = "binomial")
cl <- survival::clogit(union01 ~ wage + exper + rural + strata(id), data = UnionWage, method = "exact")
rbind(panglm = coef(f_clogit), clogit = coef(cl))
```

### Two-way fixed effects

`effect = "twoways"` adds a time fixed effect via the standard
Guimaraes & Portugal (2010) alternating-projections algorithm (exact for
unbalanced panels), currently for `family = "gaussian"` only:

```{r twoways}
f_tw <- panglm(inv ~ value + capital, data = Grunfeld, index = c("firm", "year"),
                model = "within", family = "gaussian", effect = "twoways")
p_tw <- plm(inv ~ value + capital, data = Grunfeld, index = c("firm", "year"),
            model = "within", effect = "twoways")
rbind(panglm = coef(f_tw), plm = coef(p_tw))
```

## 3. Random effects

- **gaussian**: the Swamy-Arora quasi-demeaning (feasible GLS) estimator,
  identical to `plm(model = "random")`'s default.
- **poisson**: the Poisson-Gamma (negative-binomial marginal) closed-form
  likelihood -- `pglm`'s `model = "random", family = poisson` estimator. The
  fitted dispersion parameter is the Gamma frailty shape.
- **binomial**: the individual random intercept is integrated out via
  Gauss-Hermite quadrature (`R` nodes, default 21), matching `pglm`'s
  `model = "random", family = binomial(...)`.

```{r random-poisson}
set.seed(42)
Grunfeld$u <- rep(rgamma(10, shape = 5, rate = 5), each = 20)
Grunfeld$count <- rpois(nrow(Grunfeld), lambda = Grunfeld$u * exp(1 + 0.0002 * Grunfeld$value))
f_re <- panglm(count ~ value + capital, data = Grunfeld,
                index = c("firm", "year"), model = "random", family = "poisson")
p_re <- pglm(count ~ value + capital, data = Grunfeld, index = c("firm", "year"),
             model = "random", family = poisson, method = "bfgs", print.level = 0, R = 5)
rbind(panglm = coef(f_re), pglm = coef(p_re)[1:3])
```

```{r random-binomial}
f_bin <- panglm(union ~ wage + exper + rural, data = UnionWage,
                 index = c("id", "year"), model = "random", family = "binomial", R = 15)
p_bin <- pglm(union ~ wage + exper + rural, UnionWage, index = c("id", "year"),
              family = binomial("logit"), model = "random", method = "bfgs",
              print.level = 0, R = 15)
rbind(panglm = coef(f_bin), pglm = coef(p_bin)[1:4])
```

## 4. Standard errors

Model-based (classical) SEs are always available. `vcov(fit, type = "HC1")`
and `vcov(fit, type = "cluster")` add heteroskedasticity-robust and
cluster-robust (Stata-style CR1, clustered by the panel individual by
default) sandwich covariances for `model = "pooling"`/`"within"`:

```{r robust}
sqrt(diag(vcov(f_fe)))                       # classical
sqrt(diag(vcov(f_fe, type = "cluster")))     # cluster-robust
```

Random-effects models report model-based (information-matrix) SEs only --
the individual-level correlation is already accounted for in the
likelihood, matching the convention used by `lme4`/`glmmTMB`.

You can also request a non-classical vcov directly when fitting:
`panglm(..., vcov = "cluster")`.

## 5. Hausman test

`panglm_hausman()` compares a `"within"` and a `"random"` fit of the same
formula/family, replicating `plm::phtest()`:

```{r hausman}
panglm_hausman(f_fe, panglm(inv ~ value + capital, data = Grunfeld,
                             index = c("firm", "year"), model = "random", family = "gaussian"))
```

## 6. broom / modelsummary

`tidy()` and `glance()` methods are registered for the `generics` package's
generics, so `panglm` fits work directly with `broom`- and
`modelsummary`-based table pipelines:

```{r broom}
broom::tidy(f_pool)
```

## What's not (yet) covered

- Between and random-effects `ordinal`/`tobit` models (`pglm` has these via
  Gauss-Hermite quadrature; `panglm` doesn't yet).
- Multi-way fixed effects for `poisson`/`binomial` (only gaussian).
- Cluster/HC1-robust vcov for fixed-effects binomial (classical,
  Hessian-based SEs only for now).

See the package's `plan` history and `NEWS.md` for what's planned next.
